---
title: "p8105_hw2_asb2338"
author: "Alexa Burko"
date: "2026-10-02"
output: html_document
---
## Section 0: Libraries
```{r message=F}

library(tidyverse)

```

# Problem 1

## Load and clean NYC transit dataset
```{r message=F, warning=FALSE}
transit_df =
  read_csv("data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv", na = c("", "NA", ".")) |>
  janitor::clean_names() |>
  select(line:entry, vending, ada) |>
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO"  ~ FALSE
    ))

```

This data set contains the following variables: `r colnames(transit_df)`. My data cleaning steps so far consist of making it so all blanks, "NA"s, and "."s were treated as missing data, cleaning the variable names (making them all lowercase and snake case), only keeping the variables listed in the previous sentence, and converting the variable "entry" from a character variable with the values "YES" and "NO" into a logical variable with the values "TRUE" and "FALSE". The resulting data set after cleaning has `r nrow(transit_df)` rows and `r ncol(transit_df)` columns. The data are relatively tidy. 

There are `r transit_df |> 
  distinct(station_name, line) |> 
  nrow()`  distinct stations in this data set.

`r transit_df |> 
  filter(ada==TRUE) |> 
  distinct(station_name, line) |> 
  nrow()` of the stations are ADA compliant.

The proportion of station entrances/exits that do not have vending and do allow entrance is  `r transit_df |> 
  filter(vending == "NO") |> 
  summarize(mean(entry == TRUE))`.
  
## Reformat route variables 
```{r}
transit_tidy_df = 
  transit_df |> 
  mutate(across(route1:route11, as.character)) |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name"
  ) 
```

 `r transit_tidy_df |> 
  filter(route_name == "A") |> 
  distinct(station_name, line) |> 
  nrow()`  distinct stations serve the A train.
  
  Of the  `r transit_tidy_df |> 
  filter(route_name == "A") |> 
  distinct(station_name, line) |> 
  nrow()`  distinct stations that serve the A train, `r transit_tidy_df |> 
  filter(route_name == "A") |> 
  filter(ada == "TRUE") |> 
  distinct(station_name, line) |> 
  nrow()` are ADA compliant.
  
# Problem 2

## Read in and clean Mr. Trash Wheel data set
```{r}
mr_trash_df = 
  readxl::read_excel("data/202610 Trash Wheel Collection Data.xlsx", sheet = "Mr. Trash Wheel", range = "A2:N742") |> 
  janitor::clean_names() |>
  mutate(Wheel = "Mr",
         year = as.numeric(year),
         plastic_bottles = as.numeric(plastic_bottles),
         sports_balls = as.integer(round(sports_balls, digits = 1)))
```

## Read in and clean Professor Trash Wheel data set
```{r warning=FALSE}
prof_trash_df = 
  readxl::read_excel("data/202610 Trash Wheel Collection Data.xlsx", sheet = "Professor Trash Wheel", range = "A2:M139") |> 
  janitor::clean_names() |>
  mutate(Wheel = "Professor",
         year = as.numeric(year),
         plastic_bottles = as.numeric(plastic_bottles))
```


## Read in and clean Gwynns Falls Trash Wheel data set
```{r}
gwy_trash_df = 
  readxl::read_excel("data/202610 Trash Wheel Collection Data.xlsx", sheet = "Gwynnda the Good Wheel of the W", range = "A2:L394") |> 
  janitor::clean_names() |>
  mutate(Wheel = "Gwynn",
         year = as.numeric(year),
         plastic_bottles = as.numeric(plastic_bottles))
```

## Combine Trash Wheel data sets
```{r}
trash_df = 
  bind_rows(mr_trash_df, prof_trash_df, gwy_trash_df) |> 
  relocate(Wheel)
```

The data set has `r nrow(trash_df)` observations and `r ncol(trash_df)` variables. This data set contains the following variables: `r colnames(trash_df)`. The data in this data set range from the year `r min(trash_df$year)` to the year `r max(trash_df$year)`. All together, `r trash_df |> 
  pull(weight_tons) |> 
  sum(na.rm = TRUE)` tons of trash were collected, `r trash_df |> 
  pull(plastic_bottles) |> 
  sum(na.rm = TRUE)|>
  format(scientific = FALSE)` plastic bottles were collected, and `r trash_df |> 
  pull(plastic_bags) |> 
  sum(na.rm = TRUE)|>
  format(scientific = FALSE)` plastic bags were collected. 

The total weight of trash collected by Professor Trash Wheel was `r trash_df |> 
  filter(Wheel == "Professor") |> 
  pull(weight_tons) |> 
  sum(na.rm = TRUE)` tons. 
  
Gwynnda collected 
`r trash_df |> 
  filter(Wheel == "Gwynn") |> 
  filter(month == "June") |> 
  filter(year == "2022") |> 
  pull(cigarette_butts) |> 
  sum(na.rm = TRUE) |>
  format(scientific = FALSE)` cigarette butts in June of 2022 
  
# Problem 3
## Read in and clean baseline demographics data set
```{r message=F}

base_demo_df =
  read_csv("data/data_mci/MCI_baseline.csv", 
           skip = 1,
           na = c("", "NA", ".")) |>
  janitor::clean_names() |>
  filter(age_at_onset > current_age | is.na(age_at_onset)) |>
  mutate(
    sex = case_match(
      sex,
      1 ~ "male",
      0 ~ "female"
    ),
    apoe4 = case_match(
      apoe4,
      1 ~ "carrier",
      0 ~ "non-carrier"
    )) 

```
There were many important steps in importing this data set. First, I skipped the first row of the csv file when importing it because the first line was more of a data dictionary rather than actual data. I made it so all blanks, "NA"s, and "."s were treated as missing data. I cleaned the variable names (making them all lowercase and snake case). I removed the observations where the age_at_onset was less than or equal to the current_age (meaning that I removed all the observations where the participant already had MCI at baseline). I changed the values of the variable "sex" to be equal to "male" and "female" rather than 1 and 0. Finally, I changed the values of the variable "apoe4" to be equal to "carrier" and "non-carrier" rather than 1 and 0.

The data set has `r nrow(base_demo_df)` observations and `r ncol(base_demo_df)` variables. This data set contains the following variables: `r colnames(base_demo_df)`. 

Of the `r nrow(base_demo_df)` participants that were recruited, `r sum(!is.na(base_demo_df$age_at_onset))` developed MCI at some point during the follow-up period.

The average baseline age is `r round(mean(pull(base_demo_df, current_age)),2)` years. 

The proportion of women in the study that are APOE4 carriers is  `r base_demo_df |> 
  filter(sex == "female") |> 
  summarize(mean(apoe4 == "carrier"))`.
  
## Read in and clean amyloid data set
```{r message=F, warning=FALSE}

amyloid_df =
  read_csv("data/data_mci/mci_amyloid.csv", 
           skip = 1,
           na = c("", "NA", ".")) |>
  janitor::clean_names() |>
  rename(
    id = study_id
  ) |>
  pivot_longer(
    baseline:time_8,
    names_to = "time_elapsed",
    names_prefix = "time_",
    values_to = "amyloid_ratio"
  ) |> 
  mutate(
    time_elapsed = replace(time_elapsed, time_elapsed=="baseline", "0"),
    amyloid_ratio = as.numeric(amyloid_ratio)
  )

```
In importing and cleaning this data set, I first skipped the first row of the csv file when importing it because the first line was more of a data dictionary rather than actual data. I made it so all blanks, "NA"s, and "."s were treated as missing data. I cleaned the variable names (making them all lowercase and snake case). I changed the name of the variable "study_id" to "id" so that it matches that of the baseline demographics data set. I reformatted the time variables so that the time (in years) that elapsed since the study baseline to the visit where biomarker Amyloid 42/40 ratio was measured is one variable called "time_elapsed" and the values of the biomarker Amyloid 42/40 ratio are in a new variable called "amyloid_ratio". Finally, I changed the "baseline" value of the new variable "time_elapsed" to be "0", so that it follows the same convention of the other values of that variable. 

## Compare participants in data sets
```{r}

only_in_base_df <- setdiff(base_demo_df$id, amyloid_df$id)

only_in_amyloid_df <- setdiff(amyloid_df$id, base_demo_df$id)

```
There are 8 participants that appear only in the baseline data set and not in the amyloid data set.
There are 16 participants that appear only in the amyloid data set and not in the base line data set.
There are a total of 24 participants that only appear in one data set or the other. 


## Combine data sets
```{r}
mci_df = 
  inner_join(base_demo_df, amyloid_df, by = "id")
```
After combining the baseline demographics data set and the amyloid data set, the resulting data set has `r nrow(mci_df)` observations and `r ncol(mci_df)` variables. This data set contains the following variables: `r colnames(mci_df)`. 

There are `r mci_df |> 
  distinct(id) |> 
  nrow()`  participants in this data set.
  
The average age of onset of mild cognitive impairment (MCI) is `r round(mean(pull(mci_df, age_at_onset), na.rm = TRUE),2)` years and the average Amyloid 42/40 ratio is `r round(mean(pull(mci_df, amyloid_ratio), na.rm = TRUE),2)`

## Exporting result as a CSV to data dictionary
```{r}
write_csv(mci_df, "data/data_mci/MCI.csv")
```

